---
title: "Data Preprocessing"
author: "Kathy Liu"
date: "2025-12-7"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Preprocessing

------------------------------------------------------------------------

## 1. Exploratory Data Analysis


### 1.0 Load Packages
```{r message=FALSE, warning=FALSE}
load_packages <- function(pkgs) {
  missing <- setdiff(pkgs, rownames(installed.packages()))
  if (length(missing)) install.packages(missing, dependencies = TRUE)

  pkgs_ordered <- if ("dplyr" %in% pkgs) c(setdiff(pkgs, "dplyr"), "dplyr") else pkgs

  invisible(lapply(pkgs_ordered, function(p) {
    suppressPackageStartupMessages(library(p, character.only = TRUE))
  }))

  if (!requireNamespace("conflicted", quietly = TRUE)) install.packages("conflicted")
  suppressPackageStartupMessages(library(conflicted))
  conflict_prefer("select", "dplyr")
  conflict_prefer("filter", "dplyr")
  conflict_prefer("rename", "dplyr")
  conflict_prefer("lag",    "dplyr")
  invisible(TRUE)
}

packages <- c(
  "tidyr","janitor","naniar","dplyr","skimr","magrittr","forcats","car","tibble","knitr","data.table",
  "mice",
  "ggplot2","GGally","patchwork","corrplot","ggcorrplot","scales","rpart.plot",
  "glmnet","glmtoolbox","caret","randomForest","xgboost","rpart",
  "pROC","yardstick","DescTools","probably","tableone",
  "htmlTable","htmltools","webshot"
)

load_packages(packages)

```



### 1.1 Data Overview
```{r}
# Covariates 
cov_load <- load("/Users/spectremac/Desktop/BMI706/Final_Project/covariate_df.rData")
cov_data <- get(cov_load[1])
cov_data <- as.data.frame(cov_data)

# Questionnaire (for outcome variables)
quest_data <- fread("/Users/spectremac/Desktop/BMI706/Final_Project/questionnaire.csv")|> as.data.frame()
```


### 1.2 Extract and rename variables
```{r}
# Select variables
## covariate df
cov_data <- cov_data |> 
  select(
    SEQN,
    
    # Demographics
    RIDAGEYR,   # Age in years
    RIAGENDR,   # Gender
    RIDRETH1,   # Race/Ethnicity
    DMDEDUC2,   # Education level
    INDFMPIR,   # Family income to poverty ratio

    # Blood Pressure & Anthropometry
    BPXSY1,     # Systolic blood pressure (1st reading)
    BPXSY2,     # Systolic blood pressure (2nd reading)
    BPXDI1,     # Diastolic blood pressure (1st reading)
    BPXDI2,     # Diastolic blood pressure (2nd reading)
    BMXBMI,     # Body Mass Index (kg/m²)
    BMXWAIST,   # Waist circumference (cm)
    BPXSY3,     # Systolic blood pressure (3rd reading)
    BPXDI3,     # Diastolic blood pressure (3rd reading)

    # Blood Lipids & Glycemic Markers
    LBXTC,      # Total cholesterol (mg/dL)
    LBDHDD,     # Direct HDL cholesterol (mg/dL)
    LBDLDL,     # LDL cholesterol (mg/dL)
    LBXTR,      # Triglycerides (mg/dL)
    LBXSGL,     # Glucose (serum, mg/dL)
    LBXGH,      # Glycohemoglobin (HbA1c)

    # Additional Lab Biomarkers
    LBXPLTSI,   # Platelet count
    LBXHGB,     # Hemoglobin
    LBXHCT,     # Hematocrit
    LBXSATSI,   # Iron saturation (%)
    LBXSIR,     # Serum iron
    LBXSCR,     # Creatinine
    LBXSUA,     # Uric acid
    LBXWBCSI    # White blood cell count
  )


## questionnaire df
quest_data <- quest_data |>
  select(
    SEQN,
    
    # Outcome
    DIQ010, # diabetes
    MCQ160F, # stroke
    MCQ082, # celiac disease
    MCQ160A, # arthritis
    MCQ160E, # heart attack
    MCQ160L, # liver disease
    MCQ010, # asthma
    MCQ160O, # COPD
    
    # Others
    BPQ020,    # Ever told had high blood pressure
    BPQ080,    # Currently taking BP medication
    SMD057,    # Cigarettes per day
    PAD615,    # Vigorous physical activity per day
    PAD630,    # Moderate physical activity per day
    SLD010H,   # Sleep hours
    ALQ130,    # Alcohol intake (past 12 months)
  )
```


### 1.3 Merge dataframe

```{r}
# merge datasets
raw_data <- cov_data|>
  left_join(quest_data, by = 'SEQN')
```


```{r}
#summary(raw_data)
```


### 1.4 Replace extreme, unknown, refused to answer value with NA

```{r}
# replace extreme values with NAs
raw_data <- raw_data |>
  mutate(
    SMD057 = na_if(SMD057, 999),
    PAD615 = na_if(PAD615, 9999),
    PAD630 = na_if(PAD630, 9999),
    SLD010H = na_if(SLD010H, 99),
    ALQ130 = na_if(ALQ130, 999)
  )

# replace extreme blood pressure values with NAs
raw_data <- raw_data |>
  mutate(
    BPXDI1 = na_if(BPXDI1, 0),
    BPXDI2 = na_if(BPXDI2, 0),
    BPXDI3 = na_if(BPXDI3, 0),
    BPXSY1 = na_if(BPXSY1, 0),
    BPXSY2 = na_if(BPXSY2, 0),
    BPXSY3 = na_if(BPXSY3, 0)
  )

# replace implausible HDL, LDL, and total cholesterol with NAs
raw_data <- raw_data |>
  mutate(
    LBDHDD = ifelse(
      LBDHDD < 10 | LBDHDD > 100,
      NA_real_,
      LBDHDD
    ),
    LBXTC = ifelse(
      LBXTC < 100 | LBXTC > 500,
      NA_real_,
      LBXTC
    ),
    LBDLDL = ifelse(
      LBDLDL < 30 | LBDLDL > 300,
      NA_real_,
      LBDLDL
    )
  )

# replace (77, 99, 9, 7) with NAs 
raw_data <- raw_data|>
  mutate(across(
    .cols = c(MCQ160E,
              DMDEDUC2,
              BPQ020,
              BPQ080,
              SLD010H,
              DIQ010, # diabetes 
              MCQ160F, # stroke
              MCQ082, # celiac disease
              MCQ160A, # arthritis
              MCQ160L, # liver disease
              MCQ010, # asthma
              MCQ160O), # COPD
    ~ replace(., . %in% c(77, 99, 9, 7), NA)
  ))
```


### 1.5 Convert Variable Type

```{r}
# define factor and numeric variables

factor_vars <- c(
  "MCQ160E",   # Ever told had a heart attack (1 = Yes, 2 = No)
  "DIQ010",    # Diabetes (1 = Yes, 2 = No)
  "MCQ160F",   # stroke (1 = Yes, 2 = No)
  "MCQ082",    # celiac disease (1 = Yes, 2 = No)
  "MCQ160A",   # arthritis (1 = Yes, 2 = No)
  "MCQ160E",   # heart attack (1 = Yes, 2 = No)
  "MCQ160L",   # liver disease (1 = Yes, 2 = No)
  "MCQ010",    # asthma (1 = Yes, 2 = No)
  "MCQ160O",   # COPD (1 = Yes, 2 = No)
  
  "RIAGENDR",  # Gender (1 = Male, 2 = Female)
  "RIDRETH1",  # Race/ethnicity (categorical)
  "DMDEDUC2",  # Education level (e.g., <HS, HS, College)
  "BPQ020",    # Ever told had high blood pressure (1 = Yes, 2 = No)
  "BPQ080"    # Currently taking BP meds (1 = Yes, 2 = No)
)


num_vars <- c(
  "SMD057",    # Cigarettes per day 
  "PAD615",    # Vigorous physical activity minutes 
  "PAD630",    # Moderate physical activity minutes 
  "ALQ130",     # Alcohol use past year (frequency; ordinal categorical)
  "SLD010H",   # Hours of sleep on weekday/weekend night
  "RIDAGEYR",  # Age
  "INDFMPIR",  # Income-to-poverty ratio
  "BPXSY1", "BPXSY2", "BPXSY3",  # Systolic
  "BPXDI1", "BPXDI2", "BPXDI3",  # Diastolic
  "BMXBMI",     # Body Mass Index
  "BMXWAIST",   # Waist circumference
  "LBXTC",      # Total cholesterol
  "LBDHDD",     # HDL cholesterol
  "LBDLDL",     # LDL cholesterol
  "LBXTR",      # Triglycerides
  "LBXSGL",     # Glucose (serum)
  "LBXGH",      # Glycohemoglobin
  "LBXPLTSI",   # Platelet count
  "LBXHGB",     # Hemoglobin
  "LBXHCT",     # Hematocrit
  "LBXSATSI",   # Iron saturation
  "LBXSIR",     # Serum iron
  "LBXSCR",     # Creatinine
  "LBXSUA",     # Uric acid
  "LBXWBCSI"    # White blood cell count
)

```


```{r}
# convert variable type
raw_data <- raw_data|>
  mutate(across(all_of(factor_vars), as.factor))
```


### 1.6 Remove participants with missing values for outcome variables

```{r}
# Vector of outcome variables
outcomes <- c(
  "DIQ010",   # diabetes
  "MCQ160F",  # stroke
  "MCQ082",   # celiac disease
  "MCQ160A",  # arthritis
  "MCQ160E",  # heart attack
  "MCQ160L",  # liver disease
  "MCQ010",   # asthma
  "MCQ160O"   # COPD
)

# Remove rows with ANY NA in those outcomes + reorder columns
raw_data <- raw_data |>
  filter(if_all(all_of(outcomes), ~ !is.na(.x))) |>
  select(SEQN, all_of(outcomes), everything())

dim(raw_data)

```


### 1.7 Rename variables

```{r}
# rename variable names
raw_data <- raw_data |>
  rename(
    id_seqn = SEQN,
    heart_attack_mcq160e = MCQ160E,
    diabetes_diq010 = DIQ010,
    stroke_mcq160f = MCQ160F,
    celiac_disease = MCQ082,
    arthritis = MCQ160A,
    liver_disease = MCQ160L,
    asthma = MCQ010,
    COPD = MCQ160O,
    
    age_ridageyr = RIDAGEYR,
    gender_riagendr = RIAGENDR,
    ethnicity_ridreth1 = RIDRETH1,
    education_level_dmdeduc2 = DMDEDUC2,
    income_to_poverty_ratio_indfmpir = INDFMPIR,

    systolic_bp_1_bpxsy1 = BPXSY1,
    systolic_bp_2_bpxsy2 = BPXSY2,
    systolic_bp_3_bpxsy3 = BPXSY3,

    diastolic_bp_1_bpxdi1 = BPXDI1,
    diastolic_bp_2_bpxdi2 = BPXDI2,
    diastolic_bp_3_bpxdi3 = BPXDI3,

    bmi_bmxbmi = BMXBMI,
    waist_circumference_bmxwaist = BMXWAIST,

    total_cholesterol_lbxtc = LBXTC,
    hdl_cholesterol_lbdhdd = LBDHDD,
    ldl_cholesterol_lbdldl = LBDLDL,
    triglycerides_lbxtr = LBXTR,
    serum_glucose_lbxsgl = LBXSGL,
    glycohemoglobin_lbxgh = LBXGH,

    platelet_count_lbxpltsi = LBXPLTSI,
    hemoglobin_lbxhgb = LBXHGB,
    hematocrit_lbxhct = LBXHCT,
    iron_saturation_lbxsatsi = LBXSATSI,
    serum_iron_lbxsir = LBXSIR,
    creatinine_lbxscr = LBXSCR,
    uric_acid_lbxsua = LBXSUA,
    wbc_count_lbxwbcsi = LBXWBCSI,

    told_had_high_bp_bpq020 = BPQ020,
    taking_bp_meds_bpq080 = BPQ080,

    cigarettes_per_day_smd057 = SMD057,
    vigorous_physical_activity_min_pad615 = PAD615,
    moderate_physical_activity_min_pad630 = PAD630,

    sleep_hours_sld010h = SLD010H,
    alcohol_use_frequency_alq130 = ALQ130
  )
```


```{r message=FALSE, warning=FALSE}
# rename factor variable levels
raw_data <- raw_data |>
  mutate(
    heart_attack_mcq160e = fct_recode(heart_attack_mcq160e,
                                      "Yes" = "1", "No" = "2"),
    
    diabetes_diq010 = fct_recode(diabetes_diq010,
                                      "Yes" = "1", "No" = "2"),
    
    stroke_mcq160f = fct_recode(stroke_mcq160f,
                                      "Yes" = "1", "No" = "2"),
    
    celiac_disease = fct_recode(celiac_disease,
                                      "Yes" = "1", "No" = "2"),
    
    arthritis = fct_recode(arthritis,
                                      "Yes" = "1", "No" = "2"),
    
    liver_disease = fct_recode(liver_disease,
                                      "Yes" = "1", "No" = "2"),
    
    asthma = fct_recode(asthma,
                                      "Yes" = "1", "No" = "2"),
    
    COPD = fct_recode(COPD,
                                      "Yes" = "1", "No" = "2"),

    gender_riagendr = fct_recode(gender_riagendr,
                                 "Male" = "1", "Female" = "2"),

    ethnicity_ridreth1 = fct_recode(ethnicity_ridreth1,
                                    "Mexican_American" = "1",
                                    "Other_Hispanic" = "2",
                                    "Non_Hispanic_White" = "3",
                                    "Non_Hispanic_Black" = "4",
                                    "Other_Race" = "5"),

    education_level_dmdeduc2 = fct_recode(education_level_dmdeduc2,
                                          "<9th_Grade" = "1",
                                          "9-11th_Grade" = "2",
                                          "High_School_GED" = "3",
                                          "Some_College_AA" = "4",
                                          "College_Grad_or_Above" = "5"),

    told_had_high_bp_bpq020 = fct_recode(told_had_high_bp_bpq020,
                                         "Yes" = "1", "No" = "2"),

    taking_bp_meds_bpq080 = fct_recode(taking_bp_meds_bpq080,
                                       "Yes" = "1", "No" = "2")

  )
```


```{r}
# re-order to make all factor variables be in front
factor_vars_order <- c(
  "id_seqn",
  "heart_attack_mcq160e",
  "diabetes_diq010",
  "stroke_mcq160f",
  "celiac_disease",
  "arthritis",
  "liver_disease",
  "asthma",
  "COPD",
  "age_ridageyr",
  "gender_riagendr",
  "ethnicity_ridreth1",
  "education_level_dmdeduc2",
  "told_had_high_bp_bpq020",
  "taking_bp_meds_bpq080"
)

other_vars <- setdiff(colnames(raw_data), factor_vars_order)

raw_data <- raw_data[, c(factor_vars_order, other_vars)]
```


### 1.8 Missingness examination before data imputation

```{r}
#rename some variables
raw_data_na_tbl <- miss_var_summary(raw_data)|>
  as.data.frame()|>
  rename("Missing Count" = "n_miss",
         "Variable" = "variable",
         "Missing Percentage (%)" = "pct_miss")
kable(raw_data_na_tbl, caption = 'Missing Value Summary Before Imputation')
```


```{r}
# Save dataset before splitting and imputation
#write.csv(raw_data, "/Users/spectremac/Desktop/raw_merged_data.csv")
```



### 1.9 Data Imputation: multiple imputation 


```{r message=FALSE, warning=FALSE}

# Initialize methods & predictor matrix from the raw data
imp0 <- mice::mice(raw_data, maxit = 0, printFlag = FALSE)
method <- imp0$method
pred   <- imp0$predictorMatrix

outcomes <- c(
  "heart_attack_mcq160e",
  "diabetes_diq010",
  "stroke_mcq160f",
  "celiac_disease",
  "arthritis",
  "liver_disease",
  "asthma",
  "COPD"
)

method[c("id_seqn", outcomes)] <- ""
pred[, c("id_seqn", outcomes)] <- 0


# Keep BP collinearity tweak
systolic_vars  <- c("systolic_bp_1_bpxsy1", "systolic_bp_2_bpxsy2", "systolic_bp_3_bpxsy3")
diastolic_vars <- c("diastolic_bp_1_bpxdi1","diastolic_bp_2_bpxdi2","diastolic_bp_3_bpxdi3")
for (v in systolic_vars)  pred[v, setdiff(systolic_vars,  v)] <- 0
for (v in diastolic_vars) pred[v, setdiff(diastolic_vars, v)] <- 0


imp <- mice::mice(
  raw_data,
  m               = 5,
  method          = method,
  predictorMatrix = pred,
  seed            = 123,
  maxit           = 5,
  printFlag       = FALSE
)

# Missingness summary 
imp_long_na_tbl <- naniar::miss_var_summary(mice::complete(imp, action = "long")) %>%
  as.data.frame() %>%
  dplyr::rename("Missing Count" = "n_miss",
                "Variable"      = "variable",
                "Missing Percentage (%)" = "pct_miss")
knitr::kable(imp_long_na_tbl, caption = "Missing Value Summary After Imputation")
```



### 1.11 Derived Variables using the combined-imputed data


```{r}
# Extract first completed dataset and split back
comp1 <- mice::complete(imp, action = 1) %>% as.data.frame()

```


```{r}

# Extract first imputed dataset from the whole data
imp_data <- mice::complete(imp, action = 1) |>
  as.data.frame()

# Derive variables
imp_data <- imp_data |>
  mutate(
    # Average blood pressure
    diastolic_avg = round(
      (diastolic_bp_1_bpxdi1 +
         diastolic_bp_2_bpxdi2 +
         diastolic_bp_3_bpxdi3) / 3
    ),
    systolic_avg = round(
      (systolic_bp_1_bpxsy1 +
         systolic_bp_2_bpxsy2 +
         systolic_bp_3_bpxsy3) / 3
    ),
    
    # Equate 1 min vigorous = 2 min moderate
    # and sum into a single "equivalent minutes" variable
    physical_activity_equiv = 
      vigorous_physical_activity_min_pad615 * 2 +
      moderate_physical_activity_min_pad630
  )
```


```{r}
# Rename variable names
no_missing_data <- imp_data |>
  rename(
    "ID" = id_seqn,
    "Heart_Attack" = heart_attack_mcq160e,
    "Diabetes" = diabetes_diq010,
    "Stroke" = stroke_mcq160f,
    "Celiac_Disease" = celiac_disease,
    "Arthritis" = arthritis,
    "Liver_Disease" = liver_disease,
    "Asthma" = asthma,
    "COPD" = COPD,
    
    "Age" = age_ridageyr,
    "Gender" = gender_riagendr,
    "Ethnicity" = ethnicity_ridreth1,
    "Education_Level" = education_level_dmdeduc2,
    "Income_to_Poverty_Ratio" = income_to_poverty_ratio_indfmpir,

    "BMI" = bmi_bmxbmi,
    "Waist_Circumference" = waist_circumference_bmxwaist,

    "Total_Cholesterol" = total_cholesterol_lbxtc,
    "HDL_Cholesterol" = hdl_cholesterol_lbdhdd,
    "LDL_Cholesterol" = ldl_cholesterol_lbdldl,
    "Triglycerides" = triglycerides_lbxtr,
    "Serum_Glucose" = serum_glucose_lbxsgl,
    "Glycohemoglobin" = glycohemoglobin_lbxgh,

    "Platelet_Count" = platelet_count_lbxpltsi,
    "Hemoglobin" = hemoglobin_lbxhgb,
    "Hematocrit" = hematocrit_lbxhct,
    "Iron_Saturation" = iron_saturation_lbxsatsi,
    "Serum_Iron" = serum_iron_lbxsir,
    "Creatinine" = creatinine_lbxscr,
    "Uric_Acid" = uric_acid_lbxsua,
    "WBC_Count" = wbc_count_lbxwbcsi,

    "Told_Had_High_BP" = told_had_high_bp_bpq020,
    "Taking_BP_Meds" = taking_bp_meds_bpq080,

    "Cigarettes_Per_Day" = cigarettes_per_day_smd057,

    "Sleep_Hours" = sleep_hours_sld010h,
    "Alcohol_Use_Frequency" = alcohol_use_frequency_alq130,

    "Diastolic_BP_Average" = diastolic_avg,
    "Systolic_BP_Average" = systolic_avg,
    "Physical_Activity_Equivalent_Min" = physical_activity_equiv
  )
```


```{r}
# select variables in train data
data <- no_missing_data %>%
  select(ID, Heart_Attack, Diabetes, Stroke, Celiac_Disease, Arthritis, Liver_Disease, Asthma, COPD, Age, Gender, Ethnicity, Education_Level,
    Income_to_Poverty_Ratio, BMI, Waist_Circumference, Total_Cholesterol, HDL_Cholesterol,
    LDL_Cholesterol,Triglycerides, Serum_Glucose, Glycohemoglobin, Platelet_Count,
    Hemoglobin, Hematocrit, Iron_Saturation, Serum_Iron, Creatinine,
    Uric_Acid, WBC_Count,

    Told_Had_High_BP, Taking_BP_Meds,

    Cigarettes_Per_Day, Sleep_Hours, Alcohol_Use_Frequency,

    Diastolic_BP_Average, Systolic_BP_Average, Physical_Activity_Equivalent_Min)
```



```{r}
# Save dataset
write.csv(data, "/Users/spectremac/Desktop/BMI706/Final_Project/processed_data.csv")
```



